---
title: 'STAT 400 | Group 4 | Monte Carlo Simulation: CIED Longevity'
output:
  html_document: default
---


```{r setup, include=FALSE}
library(tidyverse)     
library(knitr)         
library(kableExtra)    

set.seed(400) 

```


## Article 

Cardiac implantable electronic devices’ longevity: A novel modelling tool for estimation and comparison  

## Project Overview

### Introduction

This project replicates and extends the Monte Carlo simulation from Defaye et al. (2025) on Cardiac Implantable Electronic Device (CIED) longevity. The original study introduced a Power Consumption Index (PCI) model to standardize battery life predictions across different pacemaker models and manufacturers.

### Research Goals
1. Replicate the PCI model and Monte Carlo simulation methodology

2. Extend the model by incorporating greater physiological variability in lead impedance

3. Compare device longevity predictions under base and extended scenarios

### Power Consumption Index Model

**Model Definition**

The core PCI model from the original study is defined as:

**PCI** = t × I / C

Where:


**t** = Time in minutes

**I** = Total current drain (µA)

**C** = Battery capacity (Ah)

### Methods: 

**Core Goal:** Test impedance variability effect on PCI predictions

**Simplified Patient Model:**
- 100,000 virtual patients
- Condition (SND/AVB), HR, ventricular pacing only
- Battery, background current, lead impedance

**Intentional Simplifications:**
- Excluded atrial parameters to focus on ventricular pacing
- Set optional features to zero for clarity
- Used single impedance value (not manufacturer-specific)

**Rationale:** Isolate effect of impedance variability while maintaining model essence
**Comparison:** Base (500Ω fixed) vs Extended (N(500,100)Ω) vs High-impedance scenario (1000Ω fixed)


### PCI MODEL

```{r}

#Pacing Current Function
p_current <- function(v, pw_ms, r_ohm, hr = 60, 
                      pace_fraction = 0.3){
  # Convert Pulse width to seconds
  pw_sec  <- pw_ms / 1000
  
  # Current during pulse
  i_pulse <- v / r_ohm
  
  # Charge per pulse in Coulombs
  q       <- i_pulse * pw_sec
  
  # Convert to microcoulombs
  q_micro <- q * 1e6
  
  # Average current over 1 minute
  i_avg_microamps <- (q_micro * hr * pace_fraction)/60
  return(i_avg_microamps)
}

#PCI + Longevity Function
long <- function(i_background, i_pacing, 
                 i_options, c_mah){
  # Convert battery capacity from mAh to µAh
  c_uah      <- c_mah * 1000
  # Total current in µA
  i_total    <- i_background + i_pacing + 
    i_options
  # Power Consumption Index (PCI)
  PCI        <- i_total / c_uah
  # Longevity in hours
  long_hours <- 1 / PCI
  # Longevity in years
  long_hours / 8760
  return(list(
    i_total = i_total,
    PCI = PCI,
    years = long_years))
  }

```



### MC SIMULATION

#### 1) Simulation

```{r}
set.seed(400)
n_patients <- 100000

# Patient condition (indication)
condition <- sample(c("SND", "AVB"), n_patients, replace = TRUE, prob = c(0.6, 0.4))
hr <- pmin(pmax(rnorm(n_patients, 60, 5), 40), 90)

# Pacing percentages (as fractions 0-1)
ventricular_pacing <- ifelse(condition == "SND",
                            rnorm(n_patients, 0.29, 0.08),   # 29% ± 8% for SND
                            rnorm(n_patients, 0.90, 0.05))   # 90% ± 5% for AVB
ventricular_pacing <- pmin(pmax(ventricular_pacing, 0), 1)  # Clamp to 0-1

# Device parameters
c_mah <- pmax(rnorm(n_patients, 1000, 80), 800)        # Battery capacity
i_background <- pmax(rnorm(n_patients, 6.5, 0.7), 4)   # Background current

# Options current 
i_options <- rep(0, n_patients)  # Set to 0 to simplify

# Create patient data
sim_data <- tibble(
  id = 1:n_patients,
  condition,
  hr,
  ventricular_pacing,
  c_mah,
  i_background,
  i_options
)


```


#### 2) Base scenario (fixed impedance = 500Ω)


```{r}

sim_base <- sim_data %>%
  mutate(
    impedance = 500,
    i_pacing = p_current(2.5, 0.4, impedance, hr, ventricular_pacing),
    i_total = i_background + i_pacing + i_options,
    PCI = i_total / (c_mah * 1000),
    long_years = 1 / PCI / 8760,
    scenario = "Base (500 Ohms)"
  )

```


#### 3.1) Extended scenario (variable lead impedance)


```{r}

sim_extended <- sim_data %>%
  mutate(
    impedance = pmax(rnorm(n_patients, 500, 100), 200),
    i_pacing = p_current(2.5, 0.4, impedance, hr, ventricular_pacing),
    i_total = i_background + i_pacing + i_options,
    PCI = i_total / (c_mah * 1000),
    long_years = 1 / PCI / 8760,
    scenario = "Extended (500 Ohms ± 100 Ohms)"
  )

```


#### 3.2) 1000Ω scenario


```{r}
# 1000Ω scenario
sim_high_imp <- sim_data %>%
  mutate(
    impedance = 1000,
    i_pacing = p_current(2.5, 0.4, impedance, hr, ventricular_pacing),
    i_total = i_background + i_pacing + i_options,
    PCI = i_total / (c_mah * 1000),
    long_years = 1 / PCI / 8760,
    scenario = "High (1000 Ohms)"
  )
```


#### 4) Summarize longevity distributions (table)

```{r}

results <- bind_rows(
  select(sim_base, id, condition, long_years, scenario),
  select(sim_extended, id, condition, long_years, scenario),
  select(sim_high_imp, id, condition, long_years, scenario)

)

results$scenario <- factor(results$scenario,
  levels = c("Base (500 Ohms)", 
             "Extended (500 Ohms ± 100 Ohms)", 
             "High (1000 Ohms)"))


summary_tbl <- results %>%
  group_by(scenario) %>%
  summarise(
    mean_years = mean(long_years),
    median_years = median(long_years),
    sd_years = sd(long_years),
    q1 = quantile(long_years, 0.25),
    q3 = quantile(long_years, 0.75),
    n = n()
  )

# Summary table
summary_display <- summary_tbl %>%
  mutate(
    n = format(n, big.mark = ",", scientific = FALSE)  
  ) %>%
  rename(
    Scenario = scenario,
    Mean = mean_years,
    Median = median_years,
    SD = sd_years,
    Q1 = q1,
    Q3 = q3,
    N = n
  )

kable(summary_display, 
      digits = 2,
      align = c('l', rep('c', 6)))

```

#### 5) Statistical test (t-test) to confirm significance

```{r}
t_test_result <- t.test(
  filter(results, scenario == "Base (500 Ohms)")$long_years,
  filter(results, scenario == "Extended (500 Ohms ± 100 Ohms)")$long_years
)

t_test_high <- t.test(
  filter(results, scenario == "Base (500 Ohms)")$long_years,
  filter(results, scenario == "High (1000 Ohms)")$long_years
)


print(t_test_result)
print(t_test_high)

```


#### 6) Plot longevity distributions 

```{r}

ggplot(results, aes(x = long_years, fill = scenario)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("#3F51B5", "#FF9800", "#388E3C")) +
  labs(
    title = "CIED Longevity: Effect of Impedance on Battery Life",
    subtitle = paste("Higher impedance (1000Ω) shows clinically meaningful increase",
                     "\nNormal variability (±100Ω) has minimal practical effect"),
    x = "Longevity (years)",
    y = "Density",
    fill = "Scenario"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(color = "darkblue", face = "bold"),
    plot.subtitle = element_text(color = "gray40", size = 11),
    axis.title = element_text(color = "black")
  )

```


### DISCUSSION  

The goal of our project was to recreate the central modeling approach described in Defaye et al. (2025), which introduced the Power Consumption Index (PCI) as a standardized way for comparing cardiac implantable electronic device longevity. The original study relied on detailed manufacturer-reported current drain data to build the model, followed by a large-scale Monte Carlo Simulation of 100,000 patients. For our recreation, we reproduced the key steps in the study by generating 100,000 virtual patients with randomized physiological and device parameters, including heart rate, pacing percentages, battery capacity, background current, and lead impedance. Using these inputs, we calculated pacing current, total device current, PCI, and predicted device longevity.

Overall, we observed the following points that aligned with the original study:

#### 1. The background current dominates the total power consumption
In our simulation, patients with low pacing percentages had longevity largely determined by background current. Across all scenarios, the mean background current was approximately 6.5 µA, while mean pacing-related current ranged from ~1.5 to 2.5 µA depending on pacing burden and impedance. Consequently, background current accounted for approximately 70–80% of the total current drain in the majority of simulated patients. As a result, variations in pacing parameters had a limited impact on longevity when background current remained constant. In the original study, the majority of the PCI is driven by the background current, so our recreation aligned closely with this finding.

#### 2. The pacing burden substantially influences longevity
Our simulation shows that higher ventricular pacing percentages had higher total current and lower projected longevity. When grouping patients by condition, those with AV block (AVB) (mean ventricular pacing ~90%) exhibited a mean projected longevity that was approximately 2–3 years shorter than patients with sinus node dysfunction (SND) (mean ventricular pacing ~30%), under otherwise similar device parameters. This aligns with the original study, where pacing current was the second-largest contributor to power consumption. 

#### 3. The battery capacity alone cannot predict longevity
Devices with higher battery capacity generally exhibited longer simulated service life. However, battery capacity alone did not uniquely determine longevity. In the simulated population, patients with similar battery capacities (e.g., ~1000 Ah) demonstrated multi-year differences in projected longevity driven by differences in total current drain from background and pacing components. This demonstrates that battery size alone does not fully determine expected device lifespan, consistent with the findings of the original study.

#### 4. Lead impedance variability impacts longevity
When simulating variable lead impedance (Extended scenario), mean device longevity slightly decreased compared to the base scenario. The absolute difference in mean longevity was approximately 0.07 years (~26 days) A t-test confirmed that the difference in mean longevity was statistically significant (p < 10^(-10)), demonstrating that lead characteristics can meaningfully affect outcomes in terms of longevity. In addition, increasing lead impedance to 1000Ω resulted in a substantial shift in projected longevity, with a mean increase of approximately 1.1 years compared with the base scenario. This effect was highly statistically significant (p < 2 × 10^(-16)). This effect represents a clinically meaningful improvement in device longevity.


### CONCLUSION

Our Monte Carlo model successfully recreated key elements of the PCI-based longevity estimation introduced by Defaye et al. (2025) while extending the original framework to examine the impact of lead impedance magnitude. Despite using fewer device-specific parameters and a simplified representation of programming options, our simulated findings closely reflected the patterns reported in the original study. 
While the addition of our extended scenario ((N(500,100)Ω) provided some meaningful difference, the addition of a high-impedance (1000Ω) scenario provided an important extension to the existing model. While normal physiological variability in impedance produced some differences in longevity, higher impedance values resulted in a pronounced increase in projected device life. This finding reinforces the relevance of lead design choices in addition to programming and battery characteristics.

These findings reinforce the use of the PCI as a standardized and adaptable framework for understanding and comparing device life under varying patient and device conditions.

The alignment of our simulation and the published results support the idea that even approximate PCI-based modeling can produce meaningful insights into device longevity. As in the original study, our results highlight meaningful practical implications. Even in a simplified form, our recreation demonstrates the impact and accuracy of the PCI model and shows the value it brings to real-world scenarios. 

### REFERENCES 

Defaye, P., Boveda, S., Billuart, J.-R., Witte, K. K., & Paton, M. F. (2025). Cardiac implantable electronic devices’ longevity: A novel modelling tool for estimation and comparison. PLOS ONE, 20(9), e0333195. https://doi.org/10.1371/journal.pone.0333195

