---
title: "<span style='color:#1E4D2B;'>Monte Carlo Simulation of CIED Longevity</span>"
subtitle: "<span style='color:#C8C372;'>STAT 400: Computational Statistics — Colorado State University</span>"
author: "Group 4: Sydney, James, Waddah"
output:
  ioslides_presentation:
    css: styles.css
---

```{r setup, include=FALSE}
library(ggplot2)

# CSU brand colors
csu_green <- "#1E4D2B"
csu_gold  <- "#C8C372"

# Demo dataset for plot
set.seed(1)
n_per_group <- 50000
results <- data.frame(
  long_years = c(
    rnorm(n_per_group, mean = 15.33, sd = 2.31),
    rnorm(n_per_group, mean = 15.26, sd = 2.37)
  ),
  scenario = factor(rep(c("Base (500Ω)", "Extended (500Ω ± 100Ω)"), each = n_per_group))
)
```

## <span style="color:#1E4D2B;">Motivation</span>

* CIED (pacemaker) batteries require surgical replacement.

* Longevity prediction helps avoid too-early or too-late replacement.

* Lead impedance affects pacing current → may affect battery life.

* We use a Monte Carlo simulation to test whether impedance variability changes predicted longevity.


## <span style="color:#1E4D2B;">Simulation Inputs</span>

* 100,000 simulated patients.

* Variables included: heart rate, pacing %, battery capacity, background current, impedance.

* Two impedance conditions:

  * **Base:** fixed at 500Ω

  * **Extended:** N(500Ω, SD = 100Ω)

* Purpose: isolate the effect of impedance variability.

## <span style="color:#1E4D2B;">Problem Statement & Methodology</span>

* **Question:** Does impedance variability meaningfully change predicted longevity?
* **Steps:**
  1. Implement PCI model.
  2. Simulate patient and device parameters.
  3. Compute longevity under both impedance scenarios.
  4. Compare means and distributions.

## <span style="color:#1E4D2B;">PCI Model: Concept Overview</span>

* PCI = electrical load relative to battery capacity.
* Lower impedance → more current → shorter longevity.
* Components:

  * Pacing current
  * Background/system current
  * Optional features current

## <span style="color:#1E4D2B;">PCI Model: Pacing Current Formula</span>

```{r}
p_current <- function(v, pw_ms, r_ohm, hr = 60, pace_fraction = 0.3){
  pw_sec  <- pw_ms / 1000
  i_pulse <- v / r_ohm
  q       <- i_pulse * pw_sec
  q_micro <- q * 1e6
  (q_micro * hr * pace_fraction) / 60
}
```

## <span style="color:#1E4D2B;">PCI Model: Longevity Formula</span>

```{r}
long <- function(i_background, i_pacing, i_options, c_mah){
  c_uah      <- c_mah * 1000
  i_total    <- i_background + i_pacing + i_options
  PCI        <- i_total / c_uah
  long_hours <- 1 / PCI
  long_hours / 8760
}
```

## <span style="color:#1E4D2B;">Monte Carlo Framework (How We Tested) </span>


**For each of 100,000 patients:**

1. Calculate longevity with perfect 500Ω impedance

2. Calculate longevity with realistic variable impedance

3. Compare the two results


**This isolates** the effect of impedance variability by controlling for other patient differences


## <span style="color:#1E4D2B;">Longevity Summary Table</span>

| Scenario         | Mean  | Median | SD   | Q1    | Q3    | n       |
| ---------------- | ----- | ------ | ---- | ----- | ----- | ------- |
| Base (500Ω)      | 15.33 | 15.12  | 2.31 | 13.68 | 16.76 | 100,000 |
| Extended (±100Ω) | 15.26 | 15.08  | 2.37 | 13.58 | 16.73 | 100,000 |


* Longevity slightly lower under impedance variability.
* Differences are small → model is robust.

## <span style="color:#1E4D2B;">T-Test Results</span>

```
Welch Two Sample t-test

t = 6.61
p-value = 3.76e-11

95% CI: 0.049 to 0.090 years
Mean(Base)     = 15.33
Mean(Extended) = 15.26
```


## <span style="color:#1E4D2B;">Longevity Distribution Plot</span>

```{r, echo=FALSE, fig.height=3.5, fig.width=8}
ggplot(results, aes(x = long_years, fill = scenario)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c(csu_green, csu_gold)) +
  labs(
    title = "CIED Longevity: Base vs Extended Impedance",
    x = "Longevity (years)",
    y = "Density",
    fill = "Scenario"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(color = csu_green, face="bold"),
    axis.title = element_text(color = csu_green)
  )
```
  
  - Strong overlap → impedance variability has small effect.
  - Extended scenario shifts slightly left → minor longevity reduction.

## <span style="color:#1E4D2B;">Discussion</span>

* Background/system current has strongest influence on longevity.
* Impedance variability causes small, predictable changes.
* Pacing burden + background current dominate longevity outcomes.

## <span style="color:#1E4D2B;">Limitations & Future Work</span>

* Impedance treated as static (real devices drift).
* Model excludes multi-lead systems and adaptive algorithms.
* Could expand with dynamic impedance or clinical validation.


## <span style="color:#1E4D2B;">Conclusion</span>

* Impedance variability slightly reduces longevity.
* Predictions remain stable across realistic impedance ranges.
* Monte Carlo simulation provides strong framework for CIED evaluation.

## <span style="color:#1E4D2B;">Thank You</span>

Questions?

